---
title: "Selma_iCMS_pred_v2"
output: html_document
date: "2023-10-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(Seurat)
library(patchwork)
library(dplyr)
library(ComplexHeatmap)
library(RColorBrewer)
library(ggplot2)
library(SeuratData)
library(DESeq2)
library(gridExtra)
library(reshape2)
library('biomaRt')
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(ggdendro)
library(rlang)
library(grid)
library(gtable)
library(viridis)
library(InteractiveComplexHeatmap)
library(clipr)
```

Loading counts, metadata, scores
```{r}
# Counts, patient metadata, scores
cnts = read.delim("~/Desktop/CRC/CRC_TPM_1063_counts.csv", sep = ',', header = T, stringsAsFactors = F, row.names = "X")

Meta = read.delim("~/Desktop/CRC/Supplementary_Table_01_patient_metadata.csv", sep = ',', header = T, stringsAsFactors = F)

# Removing counts columns that don't have patient metadata 
keep = which(colnames(cnts) %in% gsub("-",".",Meta$DNA.Tumor.Sample.Barcode))
cnts = cnts[,keep]
```

Loading table of up/downregulated genes for iCMS2/3
```{r}
markers <- read.delim("~/Desktop/CRC/Selma/supp_table.csv", sep = ',', header = T, stringsAsFactors = F)
```

Create binary matrix of up/downregulated markers
```{r}

```

Checking that we have 716 unique markers
```{r}
marker_list <- unique(c(markers$iCMS2_Up,markers$iCMS2_Down,markers$iCMS3_Up,markers$iCMS3_Down))
marker_list_iCMS2 <- unique(c(markers$iCMS2_Up,markers$iCMS2_Down))
marker_list_iCMS3 <- unique(c(markers$iCMS3_Up,markers$iCMS3_Down))
```

Splitting markers by iCMS type
```{r}
iCMS2 <- unique(c(markers$iCMS2_Up,markers$iCMS2_Down))
iCMS2_df <- data.frame(markers = iCMS2, status = 'iCMS2')
iCMS3 <- unique(c(markers$iCMS3_Up,markers$iCMS3_Down))
#iCMS3 <- c(iCMS3, rep('', length(iCMS2) - length(iCMS3)))
iCMS3_df <- data.frame(markers = iCMS3,status = 'iCMS3')
i2_i3 <- rbind(iCMS2_df,iCMS3_df)
i2_i3$status <- factor(i2_i3$status)
```


####
iCMS2-3 PREDICTION - WORKS
####

IF NEEDED
```{r}
# Making a mart object with ensembl ID to hgnc symbol equivalency
library('biomaRt')
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
genes <- rownames(cnts)
g_list <- getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id","hgnc_symbol"),values=genes,mart= mart)

# Removing duplicates
dup = g_list$hgnc_symbol[which(duplicated(g_list$hgnc_symbol))]
g_list$dup = FALSE
g_list$dup[which(g_list$hgnc_symbol %in% dup)] <- TRUE
g_list$hgnc_symbol[which(g_list$dup == TRUE)] <- g_list$ensembl_gene_id[which(g_list$dup == TRUE)]
#
keep = which(rownames(cnts) %in% g_list$ensembl_gene_id )
cnts = cnts[keep,]
m = match(rownames(cnts) , g_list$ensembl_gene_id)
rownames(cnts) <- g_list$hgnc_symbol[m]
#
```

Matrix form
```{r}
cnts_mat <- as.matrix(cnts)
```

log2-transform and quantile normalize RNA-seq data
```{r}
cnts_mat <- limma::normalizeQuantiles(log2(cnts_mat+.25))
```

```{r}
# scale and center data, basically a wrapper for scale() function
library(CMScaller)
cnts_mat <- ematAdjust(cnts_mat)
```
####### Nearest template prediction
clean emat - for distCosine 0-imputation should be tested
```{r}
keepP <- stats::complete.cases(cnts_mat)
if (sum(!keepP) > 0) {
    cnts_mat <- cnts_mat[keepP,,drop = FALSE]
}
```
clean markers df
```{r}
keepT <- i2_i3$markers %in% rownames(cnts_mat)
if (sum(!keepT) > 0) {
    i2_i3 <- i2_i3[keepT,]
}
```
prepareInput 
```{r}
N <- ncol(cnts_mat)
K <- nlevels(i2_i3$status)
S <- nrow(i2_i3)
P <- nrow(cnts_mat)
```

```{r}
class.names <- levels(i2_i3$status)
i2_i3$status <- as.numeric(i2_i3$status)
```

```{r}
# provide warning if emat seems non-normalized
cnts_mat.mean <- round(mean(cnts_mat),2)
if (abs(cnts_mat.mean) >1) {
    isnorm <- " <- check feature centering!"
    cnts_mat.sd <- round(stats::sd(cnts_mat),2)
    warning(paste0("emat mean=", cnts_mat.mean, "; sd=", cnts_mat.sd, isnorm),
            call.=FALSE)
}
```
output classification overview
```{r}
feat.class <- paste(range(table(i2_i3$status)),collapse = "-")
# matching vector for cnts_mat and markers df
mm <- match(i2_i3$markers, rownames(cnts_mat),nomatch = 0)

if (!all(rownames(cnts_mat)[mm] == i2_i3$markers)) {
    stop("error matching probes, check rownames(cnts_mat) and i2_i3$markers")
}

# if features are reused across classes sample(..., replace=TRUE)
pReplace <- length(i2_i3$markers) > length(unique(i2_i3$markers))
```
prepareTemplates
```{r}
tmat <- matrix(rep(i2_i3$status,K), ncol = K) # templates matrix
for (k in seq_len(K)) tmat[,k] <- as.numeric(tmat[,k] == k)
if (K == 2) tmat[tmat==0] <- -1
```
selectDistance
```{r}
sim_fn <- function(x,y) corCosine(x,y)
simToDist <- function(cos.sim) sqrt(1/2 * (1-cos.sim))
```
ntpFunction 
```{r}
 ntp_fn <- function(n) {

        # sample-templates correlations
        n.sim <- as.vector(sim_fn(cnts_mat[mm,n, drop = FALSE],tmat))

        # optimized for speed not readability
        # matrix(emat[,n][sample.int... makes permuted matrix
        # apply(simFun... calculates correlation and return max value

        n.sim.perm.max <- apply(sim_fn(
                matrix(cnts_mat[,n][sample.int(P, S*nPerm, replace=TRUE)],
                       ncol = nPerm), tmat), 1, max)


        n.ntp <- which.max(n.sim)
        # estimate p-value
        n.sim.ranks <- rank(-c(n.sim[n.ntp],(n.sim.perm.max)))
        n.pval <- n.sim.ranks[1]/length(n.sim.ranks)

        # return results
        return(c(
            n.ntp,                # prediction
            simToDist(n.sim),     # distance to all templates
            n.pval))              # p-value
    }
```

OPTION 1
If not setting a seed and ncores  - enforce PARALLELIZED processing
```{r}
# paralellizedPrediction ##################################################
    # try parallelization if package is available and nCores not set to 1

# About nCores:
# If nCores is equal to 0 (nCores == 0 is TRUE), it means that the user hasn't specified the number of CPU cores explicitly.
#In this case, the code executes parallel::detectCores(). detectCores() is a function from the "parallel" package in R. It automatically detects the number of available CPU cores on the system and returns this value.
#If nCores is not equal to 0 (i.e., the user has specified a non-zero value), the code simply returns the user-specified value stored in nCores.

### If not setting a seed and ncores  
nPerm <- 2000
nCores=0
funVal <- vector(mode = "numeric", length = 2+K)

nCores <-  ifelse(nCores == 0,
                    parallel::detectCores(),
                    nCores)

options(mc.cores = nCores)

# avoids un-neccessary dispatches but
# memory and system.time scales with nPerm
nParts <- split(seq_len(N), cut(seq_len(N), nCores, labels=FALSE))

res3 <- parallel::mclapply(nParts, function(n)
    vapply(n, ntp_fn, funVal))
res3 <- data.frame(t(do.call(cbind, res3)))

# Prepare Output
colnames(res3) <- c("prediction",
                       paste0("d.",class.names),
                      "p.value")

res3$prediction <- factor(class.names[res3$prediction], levels = class.names)
rownames(res3) <- colnames(cnts_mat)
res3$p.value[res3$p.value < 1/nPerm] <- 1/nPerm
res3$FDR <- stats::p.adjust(res3$p.value, "fdr")

# returnOutput 
res3
```

OPTION2
If setting a seed and ncores - enforce SERIAL processing 
```{r}
# Serialized Prediction
nPerm <- 1000
set.seed(7)
nCores <- 1 # Always set ncores to 1 if seed is set

res2 <- lapply (seq_len(N), ntp_fn)
res2 <- data.frame(do.call(rbind,res2))

# Prepare Output
colnames(res2) <- c("prediction",
                       paste0("d.",class.names),
                      "p.value")

res2$prediction <- factor(class.names[res2$prediction], levels = class.names)
rownames(res2) <- colnames(cnts_mat)
res2$p.value[res2$p.value < 1/nPerm] <- 1/nPerm
res2$FDR <- stats::p.adjust(res2$p.value, "fdr")

# returnOutput 
res2
```

####
Add to Meta
####

```{r}
Meta_corrected <- Meta[Meta$RNA.Tumor.Sample.Barcode %in% rownames(res2),]
m = match(Meta_corrected$RNA.Tumor.Sample.Barcode, rownames(res2))
Meta_corrected$iCMS.pred <- res2$prediction[m]
Meta_corrected$iCMS.pval <- res2$p.value[m]
```

Add new category: undefined for pval >= 0.05
```{r}
Meta_corrected$iCMS.pred <- as.character(Meta_corrected$iCMS.pred)
Meta_corrected$iCMS.pred[Meta_corrected$iCMS.pval > 0.05] <- "Undefined"
Meta_corrected$iCMS.pred <- as.factor(Meta_corrected$iCMS.pred)
```
